---
title: "Classification Models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{classification}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette will guide you through the differential expression analysis of your data. We will load HDAnalyzeR and dplyr, load the example data and metadata that come with the package and initialize the HDAnalyzeR object.

```{r setup, results = 'hide', message = FALSE, warning = FALSE}
library(HDAnalyzeR)
library(dplyr)

hd_obj <- hd_initialize(dat = example_data, 
                        metadata = example_metadata, 
                        is_wide = FALSE, 
                        sample_id = "DAid",
                        var_name = "Assay",
                        value_name = "NPX")
```

First, we will create the data split object using the `hd_split_data()` function. This function will create a list of data frames with the training and testing sets. We can change the ratio of the training and testing sets, the seed for reproducibility, and the variable to classify. We can also add metadata columns as predictors in this stage. In this example, we will use the `Disease` column as the variable to classify and the `Sex` and `Age` columns as a metadata predictor.

```{r, message = FALSE, warning = FALSE}
split_obj <- hd_split_data(hd_obj, 
                           variable = "Disease", 
                           ratio = 0.8, 
                           seed = 123, 
                           metadata_cols = c("Sex", "Age"))
```

let's start with a regularized regression LASSO model via `hd_model_rreg()`. Exactly like in the previous vignette with the differential expression functions, we have to state the case and control(s) groups. To do specifically LASSO we will set the `mixture` parameter to 1. We will also set the `verbose` parameter to `FALSE` to not print the progress of the model in shake of clarity for this vignette.

```{r, message = FALSE, warning = FALSE}
model_res <- hd_model_rreg(split_obj,
                           case = "AML",
                           control = c("CLL", "MYEL", "GLIOM"),
                           grid_size = 5,
                           mixture = 1,
                           verbose = FALSE)

model_res$final_workflow
model_res$metrics
model_res$roc_curve
model_res$probability_plot
model_res$var_imp_plot
```

We can change several parameters in the `hd_model_rreg()` function. For example, we can change the number of cross-validation folds, the number of grid points for the hyperparameter optimization, or the feature correlation threshold. Also, exactly as with the DE functions, if the `control` parameter is not set, the function will use all the other classes as controls. For more information, please refer to `hd_model_rreg()` documentation.

We will also set mixture to NULL to allow the model to optimize this parameter (elastic net regression instead of LASSO regression) and set a palette for our classes.

```{r, message = FALSE, warning = FALSE}
model_res <- hd_model_rreg(split_obj,
                           case = "AML",
                           cv_sets = 3,
                           grid_size = 5,
                           cor_threshold = 0.7,
                           palette = "cancers12",
                           verbose = FALSE)

model_res$final_workflow
model_res$roc_curve
model_res$probability_plot
model_res$var_imp_plot
```

We can use a different variable to classify like `Sex` and even a different algorithm like random forest via `hd_model_rf()`. However, do not forget that we should create a new split object for this new model. In this case, because the classes are already balanced, we will not use the `balance_groups` parameter to FALSE to consider all the samples in the training dataset. Let's also remove everything except from number of features and AUC from the variable importance plot title.

```{r, message = FALSE, warning = FALSE}
split_obj <- hd_split_data(hd_obj, variable = "Sex", ratio = 0.8)
                               
model_res <- hd_model_rf(split_obj,
                       variable = "Sex",
                       case = "F",
                       palette = "sex",
                       cv_sets = 3,
                       grid_size = 5,
                       balance_groups = FALSE,
                       plot_title = c("features", "auc"),
                       verbose = FALSE)

model_res$final_workflow
model_res$roc_curve
model_res$probability_plot
model_res$var_imp_plot
```

If our data have a single predictor, we can use `hd_model_lr()` instead of `hd_model_rreg()` to perform a logistic regression. Random forest can be used as it was for multiple predictors.

```{r, message = FALSE, warning = FALSE}
hd_obj_single <- hd_initialize(dat = example_data |> filter(Assay == "ADA"), 
                               metadata = example_metadata, 
                               is_wide = FALSE, 
                               sample_id = "DAid",
                               var_name = "Assay",
                               value_name = "NPX")

split_obj <- hd_split_data(hd_obj_single, variable = "Disease", ratio = 0.8)

model_res <- hd_model_lr(split_obj, case = "AML", palette = "cancers12", verbose = FALSE)

model_res$roc_curve
model_res$probability_plot
```

At this point we should also check how our selected protein features look in boxplots. We will run a model as before, extract the features, select the top-9 of them based on their importance in the model and plot them with `hd_plot_feature_boxplot()`. We can either plot case vs control or case vs all other classes by changing the `type` argument.

> 📓 In case you have metadata variables as features, you will have to remove them from the feature vector before using the `hd_plot_feature_boxplot()` function as it is made to visualize protein features.

```{r, message = FALSE, warning = FALSE}
hd_obj <- hd_initialize(dat = example_data, 
                        metadata = example_metadata, 
                        is_wide = FALSE, 
                        sample_id = "DAid",
                        var_name = "Assay",
                        value_name = "NPX")

split_obj <- hd_split_data(hd_obj, variable = "Disease", ratio = 0.8)

model_res <- hd_model_rreg(split_obj, case = "AML", cv_sets = 3, grid_size = 5, verbose = FALSE)

features <- model_res$features |> arrange(desc(Scaled_Importance)) |> head(9) |> pull(Feature)

hd_plot_feature_boxplot(hd_obj, 
                        features = features, 
                        case = "AML", 
                        palette = "cancers12", 
                        type = "case_vs_control")

hd_plot_feature_boxplot(hd_obj, 
                        features = features, 
                        case = "AML", 
                        palette = "cancers12", 
                        type = "case_vs_all")
```

We can also do multiclassification predictions with all available classes in the data. The only thing that we should change is set the `case` argument to NULL so that the model understands that we want to classify all the classes. Let's see an example with regularized regression!

```{r, message = FALSE, warning = FALSE}
model_res <- hd_model_rreg(split_obj, 
                           case = NULL, 
                           cv_sets = 3, 
                           grid_size = 5, 
                           palette = "cancers12",
                           verbose = FALSE)

model_res$final_workflow
model_res$roc_curve
model_res$probability_plot
model_res$var_imp_plot
```

To summarize the results for multiple binary models we can use the `hd_plot_model_summary()` function. We can create models of different cases and compare them. Let's run three different models for three different cancers and summarize them.

> 📓 Do not forget that Ovarian Cancer is sex specific and we should consider run the analysis only with samples of that sex. We can easily integrate that into our pipeline using the `hd_filter_by_sex()` function.

```{r, message = FALSE, warning = FALSE}
model_aml <- hd_model_rreg(split_obj, case = "AML", cv_sets = 3, grid_size = 5, verbose = FALSE)

model_gliom <- hd_model_rreg(split_obj, case = "GLIOM", cv_sets = 3, grid_size = 5, verbose = FALSE)

split_obj_sex <- hd_split_data(hd_obj |> hd_filter_by_sex(variable = "Sex", sex = "F"),
                               variable = "Disease",
                               ratio = 0.8)

model_ovc <- hd_model_rreg(split_obj_sex, case = "OVC", cv_sets = 3, grid_size = 5, verbose = FALSE)
```

```{r, results = 'hide', message = FALSE, warning = FALSE}
model_summary_res <- hd_plot_model_summary(list("AML" = model_aml, 
                                                "GLIOM" = model_gliom, 
                                                "OVC" = model_ovc), 
                                           class_palette = "cancers12")
```

```{r}
model_summary_res$metrics_barplot
model_summary_res$features_barplot
model_summary_res$upset_plot_features
```

In case we have one case and multiple controls we can use the `hd_plot_feature_heatmap()` function to visualize the protein features in a heatmap. This function is usefull as we can easily see if the same features are important in multiple models. Let's see an example with the AML model and 3 different controls groups. We will also combine DE results of the same comparisons.

```{r, message = FALSE, warning = FALSE}
model_cll <- hd_model_rreg(split_obj, case = "AML", control = "CLL", cv_sets = 3, grid_size = 5, verbose = FALSE)

model_blood <- hd_model_rreg(split_obj, 
                             case = "AML", 
                             control = c("CLL", "MYEL", "LYMPH"), 
                             cv_sets = 3, 
                             grid_size = 5, 
                             verbose = FALSE)

model_all <- hd_rreg(split_obj, case = "AML", cv_sets = 3, grid_size = 5, verbose = FALSE)

de_cll <- hd_de_limma(hd_obj, case = "AML", control = "CLL", correct = c("Sex", "Age"))

de_blood <- hd_de_limma(hd_obj, 
                              case = "AML", 
                              control = c("CLL", "MYEL", "LYMPH"), 
                              correct = c("Sex", "Age"))

de_all <- hd_de_limma(hd_obj, case = "AML", correct = c("Sex", "Age"))
```

```{r, message = FALSE, warning = FALSE}
hd_plot_feature_heatmap(de_results = list("CLL" = de_cll, 
                                          "Blood" = de_blood, 
                                          "All" = de_all), 
                        model_results = list("CLL" = model_cll, 
                                             "Blood" = model_blood, 
                                             "All" = model_all), 
                        order_by = "CLL")
```

Finally, we can use the `hd_plot_feature_network()` function to visualize the protein features in a network. This function is useful as we can easily see the connections between the features and the importance of each feature in the model. Let's see an example with the same 3 models from before.

```{r, message = FALSE, warning = FALSE}
feature_panel <- model_aml[["features"]] |>
  filter(Scaled_Importance > 0.5) |>
  mutate(Class = "AML") |>
  bind_rows(model_gliom[["features"]] |>
              filter(Scaled_Importance > 0.5) |>
              mutate(Class = "GLIOM"),
            model_ovc[["features"]] |>
              filter(Scaled_Importance > 0.5) |>
              mutate(Class = "OVC"))

print(head(feature_panel))  # Preview of the feature panel

hd_plot_feature_network(feature_panel,
                        plot_color = "Scaled_Importance",
                        class_palette = "cancers12")
```

> 📓 Remember that these data are a dummy-dataset with artificial data and the results in this guide should not be interpreted as real results. The purpose of this vignette is to show you how to use the package and its functions.
