---
title: "Post Analysis: Pathway Enrichment & Automated Literature Search"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{post_analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette will guide you through the post analysis of the results obtained from the HDAnalyzeR pipeline. The post analysis consists of two possible steps: pathway enrichment analysis and automated literature search. The pathway enrichment analysis is performed using the Gene Ontology, KEGG and Reactome databases from `clusterProfiler` and `ReactomePA` packages respectively. The automated literature search is performed using the the PubMed database.

## Loading the Data

We will load HDAnalyzeR and dplyr, load the example data and metadata that come with the package and initialize the HDAnalyzeR object.

```{r setup, message = FALSE, warning = FALSE}
library(HDAnalyzeR)
library(dplyr)

hd_obj <- hd_initialize(dat = example_data, 
                        metadata = example_metadata, 
                        is_wide = FALSE, 
                        sample_id = "DAid",
                        var_name = "Assay",
                        value_name = "NPX")
```

For the Over Representation Analysis we are going to use a list of differentially expressed proteins. In this example we are going to use the up-regulated proteins. We could also use the features list from the classification models or even run both and get the intersect as it is done in the Get Started guide.

```{r, message = FALSE, warning = FALSE}
de_res <- hd_de_limma(hd_obj, case = "AML")
```

## Over Representation Analysis

First, we will perform an Over Representation Analysis (ORA) using the Gene Ontology database and the BP ontology. We will use the `hd_ora()` and `hd_plot_ora()` functions to run the analysis and plot the results respectively.

```{r, message = FALSE, warning = FALSE}
proteins <- de_res$de_res |> filter(logFC > 0 & adj.P.Val < 0.05) |> pull(Feature)

enrichment <- hd_ora(proteins, database = "GO", ontology = "BP")

enrichment_plots <- hd_plot_ora(enrichment)

enrichment_plots$dotplot
enrichment_plots$treeplot
enrichment_plots$cnetplot
```

Let's change the database and the p-value threshold.

```{r, message = FALSE, warning = FALSE}
enrichment  <- hd_ora(proteins, database = "Reactome", pval_lim_enrichment = 0.2)

enrichment_plots <- hd_plot_ora(enrichment)

enrichment_plots$dotplot
enrichment_plots$treeplot
enrichment_plots$cnetplot
```

## Gene Set Enrichment Analysis

We can also run a Gene Set Enrichment Analysis (GSEA) using the `hd_gsea()` and `hd_plot_gsea` functions. In `hd_gsea()` we can select if we want to include only up, down or both up and down regulated proteins, as well as the adjusted P value and logFC limits to consider a protein as differentially expressed. The `hd_plot_gsea()` function will plot the results.

> ⚠️ In this case, the function requires strictly differential expression results.

```{r, message = FALSE, warning = FALSE}
enrichment <- hd_gsea(de_res, database = "GO", ontology = "BP", pval_lim_enrichment = 0.5)

enrichment_plots <- hd_plot_gsea(enrichment)

enrichment_plots$dotplot
enrichment_plots$gseaplot
enrichment_plots$cnetplot
enrichment_plots$ridgeplot
```

## Searching PubMed for our Biomarkers

Finally, let's perform an automated literature search using the `hd_literature_search()`. The function requires a list with disease names as names and genes/proteins as values. We will create the list, run the search and preview the results.

```{r}
biomarkers <- list("acute myeloid leukemia" = c("FLT3", "EPO"),
                   "chronic lymphocytic leukemia" = c("PARP1", "FCER2"))

lit_res <- hd_literature_search(biomarkers, max_articles = 5)

lit_res$`acute myeloid leukemia`$FLT3$title
```

> 📓 Remember that these data are a dummy-dataset with artificial data and the results in this guide should not be interpreted as real results. The purpose of this vignette is to show you how to use the package and its functions.
