---
title: "blood_cancers_olink_case_study"
output: html_document
---

# Case Study 1: Plasma Proteomics of Hematological Malignancies Using Proximity Extension Assay

## Set-up & Data

Data is downloaded from: https://www.ebi.ac.uk/biostudies/studies/S-BSST935 

Download them and place them in the `data/` folder.

Reference: Álvez, M.B., Edfors, F., von Feilitzen, K. et al. Next generation pan-cancer blood proteome profiling using proximity extension assay. Nat Commun 14, 4308 (2023). https://doi.org/10.1038/s41467-023-39765-y

```{r, error=FALSE, warning=FALSE, message=FALSE}
library(HDAnalyzeR)
library(dplyr)
library(foreach)
library(doParallel)

cancer_dat <- hd_import_data("data/pancancer_olink_data_biostudies_v2.txt")
```

Keep only the blood cancer data.

-   Acute Myeloid Leukemia (AML)

-   Chronic Lymphocytic Leukemia (CLL)

-   Myeloma (MYEL)

Prepare metadata and HDAnalyzeR object.

```{r, error=FALSE, warning=FALSE, message=FALSE}
blood_cancers <- c("AML", "CLL", "MYEL")
metadata <- cancer_dat |> 
  select(Sample_ID, Cancer) |> 
  filter(Cancer %in% blood_cancers) |> 
  distinct()

hd_obj <- hd_initialize(
  cancer_dat |> filter(Cancer %in% blood_cancers),
  metadata,
  sample_id = "Sample_ID"
)
```

Split data to training and validation sets. We will run the whole analysis on the training set and then only evaluate the trained model on the validation set. This is done to ensure there is no data leakage.

```{r, error=FALSE, warning=FALSE, message=FALSE}
split_obj <- hd_split_data(
    hd_obj,
    variable = "Cancer",
    ratio = 0.8,
    seed = 500
)

train_obj <- split_obj$train |> 
  select(-Cancer) |> 
  hd_initialize(
    metadata = metadata |> filter(Sample_ID %in% split_obj$train$Sample_ID),
    sample_id = "Sample_ID",
    is_wide = TRUE
  )

test_obj <- split_obj$test |> 
  select(-Cancer) |> 
  hd_initialize(
    metadata = metadata |> filter(Sample_ID %in% split_obj$test$Sample_ID),
    sample_id = "Sample_ID",
    is_wide = TRUE
  )
```

## WGCNA

We will run WGCNA on the training set to identify modules of highly correlated proteins. We will also see which modules correlate better with the cancer type. To make things quicker we will use parallel processing.

```{r, error=FALSE, warning=FALSE, message=FALSE, results=FALSE}
# Set-up parallel processing
n_cores <- parallel::detectCores() - 1  # leave one core free
cl <- makeCluster(n_cores)
registerDoParallel(cl)

wgcna_res <- hd_wgcna(train_obj)
wgcna_res <- hd_plot_wgcna(train_obj, 
                           wgcna = wgcna_res, 
                           clinical_vars = "Cancer")
```

```{r}
wgcna_res$dendrogram
wgcna_res$me_cor_heatmap
```

We will then select modules that correlate more than 0.7 with CLL to run an Over-Representation Analysis (ORA) 
to identify enriched biological processes.

```{r, error=FALSE, warning=FALSE, message=FALSE}
colors <- wgcna_res$wgcna$colors

# Get the proteins of the black and green modules
green_proteins <- names(colors[colors == "green"])

cat("Number of highly correlated module proteins:", length(green_proteins))

# Run ORA with these proteins
ora_res_black <- hd_ora(green_proteins,
                        database = "GO",
                        ontology = "ALL",
                        background = "olink_explore_1536") |>  # Set background to the Olink Explore 1536 panel
  hd_plot_ora()
```

```{r}
ora_res_black$dotplot
```

## DEA

We will select only the proteins of the selected module and run DEA for CLL vs the rest. We will also plot results 
as a volcano plot.

```{r, error=FALSE, warning=FALSE, message=FALSE}
train_obj$data <- train_obj$data |> 
  select(Sample_ID, all_of(green_proteins))

de_res <- hd_de_limma(train_obj, variable = "Cancer", case = "CLL") |> 
  hd_plot_volcano(pval_lim = 0.005, 
                  logfc_lim = 2, 
                  report_nproteins = FALSE)
```

```{r}
de_res$volcano_plot
```

## Feature Selection

We will define our final biomarker panel by filtering the proteins with p-value < 0.005 and then ranking them 
based on |kME| × |log₂FC| and keep only the top 10 proteins. With this ranking we want to prioritize proteins that 
are both highly correlated with the module (hubs) and have a high fold change between CLL and the rest.

```{r, error=FALSE, warning=FALSE, message=FALSE}
# Calculate the kME
MEs_green <- wgcna_res$wgcna$MEs |> 
  select(MEgreen)

green_proteins_expr <- train_obj$data |> 
  select(Sample_ID, all_of(green_proteins)) |>
  tibble::column_to_rownames("Sample_ID")

kME <- as.data.frame(cor(green_proteins_expr, MEs_green, use = "p"))

# Combine kME and log2FC
de_res_filtered <- de_res$de_res |> 
  left_join(kME |> tibble::rownames_to_column("Feature"), by = "Feature") |> 
  filter(adj.P.Val < 0.005) |>  # Filter by p-value
  mutate(kME = abs(MEgreen)) |> 
  mutate(logFC = abs(logFC)) |> 
  mutate(rank = kME * logFC) |>  # Rank by kME * logFC
  arrange(desc(rank))

top_proteins <- de_res_filtered |> 
  slice_head(n = 10) |> 
  pull(Feature)
```

## Classification model

We will use the top 10 proteins to train a classification model. We will use a Regularized Regression model 
(LASSO) for this purpose. We will use 5-fold cross-validation to tune the model and then evaluate it on the 
held-out test set.

```{r}, error=FALSE, warning=FALSE, message=FALSE
# Prepare updated data split object by keeping only the top proteins
model_train <- split_obj$train |> 
  select(Sample_ID, Cancer, all_of(top_proteins))
model_test <- split_obj$test |> 
  select(Sample_ID, Cancer, all_of(top_proteins))
model_obj <- list("train_data" = model_train, "test_data" = model_test)
class(model_obj) <- "hd_model"

model_res <- hd_model_rreg(
  model_obj,
  variable = "Cancer",
  case = "CLL",
  grid_size = 30,
  mixture = 1,  # LASSO
  palette = "cancers12",
  plot_y_labels = TRUE,
  verbose = FALSE,
  seed = 500
)
```

```{r}
model_res$feat_imp_plot
model_res$probability_plot
model_res$roc_curve
```

## Visualize Features

We will visualize our top 4 features as boxplots.

```{r, error=FALSE, warning=FALSE, message=FALSE}
imp_proteins <- as.character(model_res$features |>
  head(4) |>
  pull(Feature))

boxplots <- hd_plot_feature_boxplot(hd_obj, variable = "Cancer", 
                        features = imp_proteins, 
                        case = "CLL",
                        palette = "cancers12") + 
                        ggplot2::facet_wrap(~Features, nrow = 1)
```

```{r}
boxplots
```

## Search in PubMed

Finally, we will search in PubMed for the top 4 proteins. As it queries a database results might change over time.

```{r, error=FALSE, warning=FALSE, message=FALSE}
features_list <- list("Chronic lymphocytic leukemia" = imp_proteins)

articles <- hd_literature_search(
  features_list, 
  max_articles = 3,
  keywords = "human"
)
```